---
title: IABA Bootcamp
date: 30 July 2019
author: Brian A. Fannin ACAS CSPA
output: 
    revealjs::revealjs_presentation:
      center: false
      css: css/custom.css
      self_contained: true
      theme: white
      transition: slide
      df_print: paged
---

```{r include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE
  , collapse = TRUE
  , error = TRUE
  , message = FALSE
  , warning = FALSE
  , fig.height = 4
)
```

# Welcome

## Agenda

* Data structure
* Data exploration
* Data visualization
* Data modeling

## Tools context

* R 
* Excel
* SQL
* Others

## %>% 

* Passes the object to the left as the first argument of the function to the right
* Enables nested functions to be read left to right, top to bottom, not inside out

## %>%

```{r message = FALSE}
library(magrittr)

exp(1)

1 %>% exp()

1 %>% 
  exp() %>% 
  log()


log(exp(1))
```

## 

* **Univariate structure**
* Univariate exploration
* Univariate visualization
* Univariate modeling

# Univariate structure

## All data is _typed_

* Continuous, double, $\mathbb{R}$
* Integer, $\mathbb{N}$
* Character
* Date

```{r}
5 + 'Steve'
```

## Vectors

* Set of values of the same data type
* Easy mapping to concepts in linear algebra
* Matrices, and higher-order arrays/tensors

$$Y = \beta_0 + \beta_1 * X$$

This:

```{r eval = FALSE}
y = b_0 + b_1 * x
```

Rather than: 

```
for i = 1 to N
  y[i] = b_0 + b_1 * x[i]
next i
```

## Excel

Vectors do not really exist in a spreadsheet

In VBA, these are arrays

```
dim claim_amount(1 to num_claims) as double
```

## SQL

Does not really exist. Individual columns may be regarded as vectors. More on SQL tables soon.

```
SELECT 
  claim_amount
FROM
  tbl_claim
```

## R

Core concept in R

```{r}
set.seed(1234)
num_claims <- 500
mean <- 10e3
cv <- 0.3
claim_amount <- rgamma(num_claims, 1 / cv^2, scale = mean * cv ^ 2)

claim_amount %>% head()
```

## Vector access

```{r collapse = FALSE}
claim_amount[1:5]
```

```{r collapse=FALSE}
claim_amount[claim_amount > 10e3]
```

## Type conversion

* Data may be converted from one type to another
* Common when converting character dates and currencies into numeric values

## 

* Univariate structure
* **Univariate exploration**
* Univariate visualization
* Univariate modeling

# Univariate exploration

## Basic 

* N -> 1 or N -> M mapping functions, basic statistics
* max
* min
* quantiles
* extremes
* 2nd and higher statistical moments

## Basic EDA

```{r}
claim_amount %>% 
  summary()

claim_amount %>% 
  var()

sd(claim_amount) / mean(claim_amount)
```

## Limited expected value

```{r}
library(actuar)

claim_lev <- claim_amount %>% 
  elev()

claim_lev(10e3)
```

## 

* Univariate structure
* Univariate exploration
* **Univariate visualization**
* Univariate modeling

# Univariate visualization

## Histogram

```{r}
hist(claim_amount)
```

## Boxplot

```{r}
boxplot(claim_amount)
```

## 

* Univariate structure
* Univariate exploration
* Univariate visualization
* **Univariate modeling**

# Univariate modeling

## Single parametric distribution

```{r}
library(MASS)

fit_gamma <- fitdistr(claim_amount, 'gamma')
fit_gamma

fit_lognorm <- fitdistr(claim_amount, 'log-normal')
fit_lognorm
```

## Maximum likelihood in Excel

* This may be done in Excel
* Form the likelihood function and use Solver
* Standard error of parameters must be coded by hand

## Inferences

```{r}
plnorm(10e3, fit_lognorm$estimate[1], fit_lognorm$estimate[2])
pgamma(10e3, fit_gamma$estimate[1], fit_gamma$estimate[2])
```

## Parametric limited expected value

```{r}
claim_lev(10e3)

levgamma(10e3, fit_gamma$estimate[1], fit_gamma$estimate[2])
```

## 

* **Multivariate structure**
* Multivariate exploration
* Multivariate visualization
* Multivariate modeling

# Multivariate structure

```{r include = FALSE}
load('data/data.rda')
```

## The table

* Set of columns
* Each column has the same type

```{r echo = FALSE}
tbl_policy %>% 
  dplyr::select(region, effective_date, premium) %>% 
  knitr::kable(
    format.args = list(big.mark = ',')
  )
```


## Alternatives

* Graph - https://neo4j.com/
* Document - https://www.mongodb.com/
* Others: columnar store, CouchDB, Hive, Spark, etc.

## Tables in R

```{r}
library(tidyverse)
```

```{r eval = FALSE}
tbl_claim <- tibble(
  occurrence_date = as.Date('2001-01-01')
  , claim_amount = claim_amount
)
```

## Relational Database Management System - RDBMS

* Microsoft Access
* Oracle
* SQL Server
* PostgreSQL

## Tables in SQL

```
CREATE TABLE tbl_policy (
	id	INTEGER NOT NULL PRIMARY KEY,
	effective_date	TEXT,
	premium	REAL,
	years_in_operation	REAL,
	number_of_employees	INTEGER,
	num_claims	INTEGER,
	expiration_date	TEXT
);
```

## Structural integrity

* Single-table constraints
  * Uniqueness of record properties: SSN, VIN, etc.
  * Value domain restrictions: 
* Multi-table constraints
  * Parent/child table, primary/foreign key

## 

* Multivariate structure
* **Multivariate exploration**
* Multivariate visualization
* Multivariate modeling

# Multivariate exploration

## Grammar of data, one table

__Basic verbs__

| Action              | SQL statement    | R Function  |
|--------------------|--------------|-----------------|
| Column-wise subset |  SELECT `X` AS [`Y`] | `select()`    |
| Row-wise subset | WHERE   | `filter()` |
| Add additional column | SELECT `X` / `Y` AS `LossRatio` | `mutate()`  |
| Apply action to a subset  | GROUP BY  | `group_by()`  |
| Summarise | SUM(X), etc. |  `summarise()`  |
| Arrange | ORDER BY  | `arrange()` |

## SQL: declarative

* SQL statement _declares_ the desired result
* Statements contain various _clauses_, which apply different aspects of the grammar
* Clauses must appear in a specific order
* Computer decides on the order of execution

## R: imperative

* Functions are executed one at a time, in the order in which they appear
* As many functions as we like, in  any order
* Changing the order may change the outcome

## Grammar of data, one table

__Basic verbs__

| Action              | SQL statement    | R Function  |
|--------------------|--------------|-----------------|
| Column-wise subset |  SELECT `X` AS [`Y`] | `select()`    |
| Row-wise subset | WHERE   | `filter()` |
| Add additional column | SELECT `X` / `Y` AS `LossRatio` | `mutate()`  |
| Apply action to a subset  | GROUP BY  | `group_by()`  |
| Summarise | SUM(X), etc. |  `summarise()`  |
| Arrange | ORDER BY  | `arrange()` |

## Column-wise subset - SQL

```
SELECT 
  effective_date
  , expiration_date
FROM 
  tbl_policy
```

## Column-wise subset - R

```{r}
library(tidyverse)

tbl_policy %>% 
  select(effective_date, expiration_date)
```

## Column-wise subset - R

Text-based search of column names

```{r}
tbl_policy %>% 
  select(contains('date')) %>% 
  head()
```

Select everything other than excluded columns

```{r }
tbl_policy %>% 
  select(-id) %>% 
  head()
```

## Row-wise subset

* By order of appearance
* Logical subset

## Row-wise subset - SQL

```
SELECT TOP 10
FROM
  tbl_policy
```

```
SELECT * 
FROM 
  tbl_policy
LIMIT 10
```

## Row-wise subset R

```{r}
tbl_policy %>% 
  slice(1:10)
```

## Row-wise subset - SQL

```
SELECT *
FROM
  tbl_policy
WHERE
  effective_date >= '2005-01-01'
  AND effective_date <= '2005-12-31'
```

## Row-wise subset - R

```{r}
tbl_policy %>% 
  filter(
    effective_date >= as.Date('2005-01-01')
    , effective_date <= as.Date('2005-12-31')
  )
```

## Mutation - SQL

```
SELECT 
  *
  , policy_duration = date(expiration_date) - date(effective_date)
FROM
  tbl_policy
```

## Mutation - R

```{r}
tbl_policy %>% 
  mutate(
    policy_duration = expiration_date - effective_date
  ) %>% 
  select(id, policy_duration)
```

## Summarization - SQL

```
SELECT
	SUBSTR(effective_date, 1, 4) AS policy_year
  , COUNT(*) AS policy_count
  , AVG(premium) AS mean_premium
FROM
  tbl_policy
GROUP BY
	SUBSTR(effective_date, 1, 4)
```

## Summarization - R

```{r}
tbl_policy %>% 
  group_by(policy_year = lubridate::year(effective_date)) %>% 
  summarise(
    policy_count = n()
    , mean_premium = mean(premium)
  )
```


## Grammar of data, two tables

| Action              | SQL equivalent    | R Function  |
|--------------------|--------------|-----------------|
| Append rows |  UNION [ALL]   | | `bind_rows()` |
| Same value for a column | INNER JOIN | `inner_join()` |
| All records in table X and only related rows from table Y | LEFT JOIN | `left_join()` |
| Only those records from table X which don't appear in table Y | LEFT JOIN and WHERE | `anti_join()`

## Append rows - SQL

```
SELECT *
FROM
  tbl_claim
UNION ALL
  tbl_other
```
## Append rows - R

```{r}
tbl_claim %>% 
  bind_rows(
    tibble(
      policy_id = 1
      , claim_amount = 5e3
    )
  ) %>% 
  filter(
    policy_id == 1
  )
```

## Inner join - SQL

```
SELECT 
  pol.id AS policy_id
  , clm.id AS claim_id
  , premium
  , claim_amount
FROM 
  tbl_policy AS pol
  INNER JOIN tbl_claim clm ON pol.id = clm.policy_id 
```

## Inner join - R

```{r}
tbl_policy %>% 
  inner_join(tbl_claim, by = c(id = 'policy_id'))
```

## Outer join - SQL

```
SELECT *
FROM 
  tbl_policy pol RIGHT JOIN tbl_claim clm ON pol.id = clm.policy_id
```

## Outer join - R

```{r}
tbl_policy %>% 
  right_join(tbl_claim, by = c(id = 'policy_id'))
```

## Anti join - SQL

```
SELECT *
FROM 
  tbl_policy pol LEFT JOIN tbl_claim clm ON pol.id = clm.policy_id
 WHERE
	clm.id IS NULL
```

## Anti-join - R

```{r}
tbl_policy %>% 
  anti_join(tbl_claim, by = c(id = 'policy_id')) %>% 
  select(id, num_claims)
```

## 

* Multivariate structure
* Multivariate exploration
* **Multivariate visualization**
* Multivariate modeling

# Multivariate visualization

## `ggplot2`

* Very popular visualization engine
* Based on the "grammar of graphics" by Leland Wilkinson
* Requires a table for construction

## Scatter plot

```{r}
tbl_policy %>% 
  ggplot(aes(years_in_operation, num_claims)) + 
  geom_point()
```

## Scatter plot

```{r}
tbl_policy %>% 
  ggplot(aes(years_in_operation, num_claims)) + 
  geom_point() + 
  geom_smooth(method = 'lm')
```

## Scatter plot

```{r}
tbl_policy %>% 
  ggplot(aes(lubridate::year(effective_date), num_claims)) + 
  geom_point()
```

## Scatter plot

```{r}
tbl_policy %>% 
  ggplot(aes(lubridate::year(effective_date), num_claims)) + 
  geom_jitter(height = NULL)
```

## Boxplot

```{r}
tbl_policy %>%
  inner_join(tbl_claim, by = c(id = 'policy_id')) %>% 
  mutate(
    occ_year = lubridate::year(occurrence_date) %>% as.factor()
  ) %>% 
  ggplot(aes(occ_year, num_claims)) + 
  geom_boxplot()
```

## Frequency by year of occurrence

```{r}
tbl_policy %>%
  inner_join(tbl_claim, by = c(id = 'policy_id')) %>% 
  mutate(
    occ_year = lubridate::year(occurrence_date)
  ) %>% 
  ggplot(aes(occ_year, num_claims)) + 
  geom_jitter(height = NULL) + 
  geom_smooth()
```

## Frequency by date of occurrence

```{r}
tbl_policy %>%
  inner_join(tbl_claim, by = c(id = 'policy_id')) %>% 
  ggplot(aes(occurrence_date, num_claims)) + 
  geom_jitter() + 
  geom_smooth()
```

## Grouping

```{r}
tbl_policy %>% 
  ggplot(aes(num_claims)) + 
  geom_histogram(aes(fill = region), position = 'dodge')
```

## Grouping

```{r}
tbl_policy %>% 
  ggplot(aes(num_claims)) + 
  geom_histogram(aes(fill = region), position = 'dodge') + 
  facet_wrap(~ region)
```

## Grouping

```{r}
tbl_policy %>% 
  ggplot(aes(num_claims)) + 
  geom_density(aes(fill = region), alpha = 0.7)
```

## 

* Multivariate structure
* Multivariate exploration
* Multivariate visualization
* **Multivariate modeling**

# Multivariate modeling

## Linear models

```{r}
fit_one <- tbl_policy %>% 
  glm(
    formula = num_claims ~ 1 + effective_date + years_in_operation + number_of_employees
    , family = poisson()
  )
```

## Model diagnostics

```{r echo = FALSE}
fit_one %>% 
  broom::glance() %>% 
  knitr::kable()
```

## Coefficient diagnostics

```{r echo = FALSE}
fit_one %>% 
  broom::tidy() %>% 
  knitr::kable()
```

## Other coefficients

```{r}
fit_two <- tbl_policy %>% 
  glm(
    formula = num_claims ~ 1 + effective_date + years_in_operation + number_of_employees
    , family = poisson(link = 'identity')
  )
```

##

```{r echo = FALSE}
fit_two %>% 
  broom::tidy() %>% 
  knitr::kable()
```

##

```{r}
fit_three <- tbl_policy %>% 
  glm(
    formula = num_claims ~ 1 + years_in_operation + number_of_employees
    , family = poisson(link = 'identity')
  )
```

##

```{r echo = FALSE}
fit_three %>% 
  broom::tidy() %>% 
  knitr::kable()
```

## Another fit

```{r}
fit_last <- tbl_policy %>% 
  glm(
    formula = num_claims ~ 1 + region + years_in_operation + number_of_employees
    , family = poisson(link = 'identity')
  )
```

## Coefficients

```{r echo = FALSE}
fit_three %>% 
  broom::tidy() %>% 
  knitr::kable()
```

## Compare models

```{r}
tbl_compare <- tibble(
  model = c('One', 'Two', 'Three', 'Four')
  , fit_obj = list(fit_one, fit_two, fit_three, fit_last)
  , family = rep('Poisson', 4)
  , link = c('exp', rep('identity', 3))
) %>% 
  mutate(
    formula = map(fit_obj, 'formula')
    , AIC = map_dbl(fit_obj, AIC)
    , deviance = map_dbl(fit_obj, 'deviance')
  ) %>% 
  dplyr::select(
    -fit_obj
  ) %>% 
  arrange(AIC)
```

## Compare models

```{r echo = FALSE}
tbl_compare %>% 
  knitr::kable()
```


# Suggestions

## Get familiar with a database
  
* SQLite https://www.sqlite.org/index.html and PostgreSQL https://www.postgresql.org/ are free
* Learn the rules of database normalization
* Get comfortable writing SQL. Use the GUI if you get stuck, but try to write it first
* Create tables with rational constraints, then try to break them. Does your database permit claims without policyholders? Negative premium amounts?

## Learn R or Python

* Both are free and open source
* Free learning resources abound
* Statistical and predictive modeling support is much stronger than in Excel

## Get comfortable, but not reliant, on spreadsheets

* LibreOffice is free and works with Excel files
* Try building very basic models in a spreadsheet, use functions like LINEST
* Get comfortable passing data in and out of Excel

## Market yourself while you learn

* Create a [Github](https://github.com/) profile
* If sharing with employers, try to restrict public projects to work you are proud of
* Work on Kaggle competitions
* Find a local meetup group for R or Python
* Prepare a talk

# Conclusion

## Our points

* Pay close attention to data structure
* Tabular data will enable solutions for most actuarial problems
* Many practical business questions may be answered simply through data manipulation
* `ggplot2` is a strong tool for data visualization
* Think about your data in terms of models
* Compare models to each other

## Thank you!!

## Resources

* https://github.com/PirateGrunt/iaba_bootcamp
* https://pirategrunt.com/raw_book/
* [R for Data Science](https://r4ds.had.co.nz/)
* https://www.coursera.org/learn/r-programming